---
title: 'enlabel: quick start'
output:
  rmarkdown::github_document: default
vignette: "%\\VignetteIndexEntry{enlabel-vignette} %\\VignetteEncoding{UTF-8} %\\VignetteEngine{knitr::rmarkdown}\n"
editor_options:
  chunk_output_type: inline
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction

R's lackluster support for variable labels is often a source of frustration, especially when working with survey data. This is because survey data: 

+ often contains long labels (i.e. questions asked to respondents);
+ comes from environments with native support for labels such as SPSS or STATA;
+ *has to be frequently transformed from wide to long format* (this is particularly true of multiple choice questions).

`enlabel` addresses the last problem in particular - labeling variable names transformed into values in a long format dataframe. Long format data, especially in the tidyverse context, lends itself more easily to certain kinds of analyses and visualisations. However, the output of these transformations usually comprises (only) numbers and question codes and can be difficult to read, interpret, and communicate. `enlabel` provides you with the tools to quickly append meaningful, human-readable descriptions to your data.

As a bonus, enlabel also provides one quality-of-life function for labeling dataframe headers. This can be useful when you are planning to export your data to a different format (e.g. to .xlsx using the `expss` package).

### Labeling long format data

The example below presents a typical task when working with survey data - selecting and summarising a group of variables sharing a common prefix. The names of the pivoted variables (i.e. question codes) are stored in the "name" column. The data comes from an unpublished study of Polish makerspaces conducted in 2022. Codes in the name column represent different kinds of tools and values in the proportion column the share of makerspaces having a prarticular tool at their disposal.

```{r message=FALSE}
library(enlabel)
library(ggplot2)
library(dplyr)
library(tidyr)


df <- haven::read_sav("makerspaces.sav")

tools_summary <- df %>%
  select(starts_with('HW')) %>%
  pivot_longer(everything()) %>%
  count(name, value) %>%
  group_by(name) %>%
  summarise(proportion = n / sum(n), value = value) %>%
  filter(value == "Y") %>% 
  select(-value)

head(tools_summary, 20)
```

`enlabel` allows you to source the dataset for labels with `extract_labels_from_data()`. The function accepts as an argument either a path to the .sav file or to an R dataframe with a `variable.labels` attribute. Importing a .sav file with the `foreign` package or using the "Export to R" functionality in LimeSurvey will both create such objects. Unfortunately, this does not work with objects imported with the `haven` package (this means that, in this particular case, `extract_labels_from_data(df)` is not going to work).

If you are working with only one survey, there is no need to assign the output of `extract_labels_from_data()` to a variable - it is stored in cache.

```{r message=FALSE, warning=FALSE, results="hide"}
extract_labels_from_data("makerspaces.sav")
```

```{r message=FALSE}
tools_summary %>% append_labels() %>% head(20)
```

If you want to visualize the data, you might want to trim the length of the labels which by default contain both the parent-question text as well as the subquestion label. You can easily discard the parent-question text by setting the `trim` argument to `TRUE`.

```{r fig.height=7, fig.width=7}
tools_summary %>% 
  append_labels(trim = TRUE) %>% 
  ggplot(aes(proportion, reorder(labels, proportion))) + 
  geom_col() +
  scale_y_discrete(labels = scales::label_wrap(40)) +
  theme_minimal() +
  theme(axis.title.y = element_blank())
```

By default, `append_labels()` searches for the codes in the "name" column. It is designed to to fit into the dplyr select-and-pivot pipeline without the need to specify any additional arguments.

### Sourcing labels from LimeSurvey

`enlabel` gives you the option of sourcing your labels directly from LimeSurvey. To do this, you need to first connect to LimeSurvey RemoteControl API. `enlabel` uses a modified version of the `limer` package to establish connection.

The creators of `limer` no longer maintain it. An up-to-date version of the package, compatible with the latest versions of LimeSurvey, can be installed directly from my github `devtools::install_github("wjk-g/limer")`.

```{r eval=FALSE}
connect2ls("your_username", "your_password", "url_to_ls_api")
extract_labels_from_ls(survey_id)
```

`append_labels()` behaves the same regardless of the source of your labels. The only difference is that labels imported from LimeSurvey do not include the parent-question text (the `trim` argument does not apply to them).

## Labeling data and exporting tables to excel

`enlabel()` gives you the tools to quickly label your data and export marginal distributions or contingency tables (or both) to Excel. The export itself is achieved with `export_to_excel()` which is a wrapper around several functions from the `expss` package. Should you want to label and export your data to Excel the process would look as follows:

```{r eval=FALSE}
# get your data

data <- read.csv("example_data.csv")

connect2ls("username", "password", "url")
extract_labels_from_ls(survey_id) # extract_labels_from_data would also work

labs <- get_colnames(data, trim = TRUE)
labelled_data <- label_variables(data, labs)

export_to_excel(labelled_data, # a df to be used for export 

export_to_excel(labelled_data, # the df to be used for export 
                varnames = list("variable1", "variable2"), # variable names for contingency tables
                max_cat = 10, # variables with > 10 unique values will not be included in the output
                bonferroni = TRUE, # whether to apply the Bonferroni correction (default: FALSE)
                overwrite = TRUE) # whether to overwrite an existing file of the same name (default: FALSE)
```


